Notes on categorical logic and term calculi

### 2025-10-24 term-calculus/cat_logic_notes.tex
type theory update

\usepackage{stmaryrd}
\newcommand{\parr}{\mathbin{\bindnasrepma}}
% \usepackage{amssymb}
\usepackage{mathpartir}

@@

\input{../preamble}
\input{../macros}
\input{../sample.tikzstyles}
\input{../hypergraph.tikzstyles}
\input{../hypergraph.tikzdefs}

@@

@@

\begin{lemma}[Substitution is associative and interchanging]
  \begin{enumerate}
  \item If $\Gamma \vdash M : A$ and $\Delta, x : A, \Delta' \vdash N : B$ and $\Psi, y : B, \Psi' \vdash P : C$, then
  \[
  (P[M/x])[N/y] = P[M[N/y]/x]
  \]
  \item If $\Gamma \vdash M : A$ and $\Delta \vdash N : B$ and $\Psi, x : A, \Psi', y : B, \Psi'' \vdash P : C$, then
  \[
  (P[M/x][N/y]) = P[N/y][M/x]
  \]
  \end{enumerate}
\end{lemma}

@@

\begin{lemma}[Substitution is admissible]
  If $\Gamma \vdash M_1 + \ldots + M_m : A$ and $\Delta, x : A, \Phi \vdash N$ are derivable, then $\Delta, \Gamma, \Phi \vdash N[M/x]$ is derivable.
\end{lemma}
\begin{proof}
  Since the substitution is admissible in the original type theory, we only need to consider the extra rule that we introduced.
  Suppose $N \Coloneqq N_1 + \ldots + N_n$ and therefore comes from the rule
  \[
  \inferrule[]{\inferrule[]{\vdots}{\Gamma \vdash M} \quad \inferrule[]{\Delta, x : A, \Phi \vdash N_1, \ldots, \Delta, x : A, \Phi \vdash N_n}{\Delta, x : A, \Phi \vdash N}}{\;}
  \]
  By IH we have derivations
  \[
    \Delta, \Gamma, \Phi \vdash N_1[M_1 + \ldots + M_m /x], \ldots, \Delta, \Gamma, \Phi \vdash N_n[M_1 + \ldots + M_m /x]
  \]
  or,
  \[
    \Delta, \Gamma, \Phi \vdash N_1[M_1/ x] + \ldots + N_1[M_m /x], \ldots, \Delta, \Gamma, \Phi \vdash N_n[M_1 / x] + \ldots + N_n[M_m /x]
  \]
  and therefore we can get a derivation 
  \[
  \Delta, \Gamma, \Phi \vdash (N_1[M_1/ x] + \ldots + N_1[M_m /x]) + \ldots + (N_n[M_1/ x] + \ldots + N_n[M_m /x])
  \]

@@

Consider the revised rules (forget about the symmetry for a moment) in \autoref{fig:slat-typetheory}.
\begin{figure}
\begin{gather*}
  % \inferrule*[right=$id$]{\vdash A\;\text{type}}{x : A \vdash x + \ldots + x : A}\\
  \inferrule*[right=$id$]{\vdash A\;\text{type}}{x : A}\\
  \inferrule*[right=$f$]{f \in \mathcal{G}(A_1,\ldots,A_n;B) \quad \Gamma_1 \vdash \sum_{j=1}^{i_1} M_{1j} : A_1 \ldots \Gamma_n \vdash \sum_{j=1}^{i_n} M_{nj} : A_n}{\Gamma_1, \ldots, \Gamma_n \vdash \sum_{k_1 = 1}^{j_1} \ldots \sum_{k_n=1}^{i_n} f(M_{1 k_1}, \ldots, M_{n k_n}) : B}\\
  \inferrule*[right=$\otimes I$]{\Gamma \vdash \sum_{i=1}^{m}M_i : A \quad \Delta \vdash \sum_{i=1}^{n} N_i : B}{\Gamma, \Delta \vdash \sum_{i=1}^{m}\sum_{j=1}^{n}\{M_i, M_j\} : A \otimes B}\\
  \inferrule*[right=$\otimes E$]{\Gamma \vdash \sum_{i=1}^{m}M_i : A \otimes B \qquad \Delta_1, x : A, y : B, \Delta_2 \vdash \sum_{i=1}^{n}N_i : C}{\Delta_1, \Gamma, \Delta_2 \vdash \sum_{1}^{i=m}\sum_{j=1}^{n}\text{match}_{A \otimes B}(M_i, xy. N_j) : C}\\
  \inferrule*[right=$\mathbf{1}I$]{\;}{() \vdash \star : \mathbf{1}}\\
  \inferrule*[right=$\mathbf{1}E$]{\Gamma \vdash \sum_{i=1}^{m} M_i : \mathbf{1} \qquad \Delta_1,\Delta_2 \vdash \sum_{i=1}^{n} N_i : A}{\Delta_1,\Gamma,\Delta_2 \vdash \sum_{i=1}^{m}\sum_{j=1}^{n} \text{match}_{\mathbf{1}}(M_i,N_j) : A}\\
  \inferrule*[right=$+I$]{\Gamma \vdash \sum_{i=1}^{i_1} M_{1i} \quad \ldots \quad \Gamma \vdash \sum_{i=1}^{i_m} M_{mi}}{\Gamma \vdash \sum_{i=1}^{m}\sum_{j=1}^{i_j} M_{ij}}
\end{gather*}
\caption{Type theory for semilattice-enriched monoidal categories}
\label{fig:slat-typetheory}
\end{figure}

@@

\begin{lemma}[Substitution is admissible]
  If $\Gamma \vdash M_1 + \ldots + M_m : A$ and $\Delta, x : A, \Phi \vdash N : B$ are derivable, then $\Delta, \Gamma, \Phi \vdash N[M/x]$ is derivable.
\end{lemma}
\begin{proof}
  \begin{enumerate}
    \item Suppose $N \Coloneqq x : A \vdash x : A$, then $A = B$ and we can just use $M_1 + \ldots + M_m$ which is $x[M_1 + \ldots + M_n / x] = x[M_1/x] + \ldots + x[M_m / x] = M_1 + \ldots + M_m$
    \item Suppose $N$ is the conclusion of the rule
    \[
      \inferrule*[right=$fI$]{f \in \mathcal{G}(A_1,\ldots,A_n;B) \quad \Gamma_1 \vdash \sum_{j=1}^{i_1} M_{1j} : A_1 \ldots \Gamma_n \vdash \sum_{j=1}^{i_n} M_{nj} : A_n}{\Gamma_1, \ldots, \Gamma_n \vdash \sum_{k_1 = 1}^{j_1} \ldots \sum_{k_n=1}^{i_n} f(M_{1 k_1}, \ldots, M_{n k_n}) : B}
    \] 
    then there exists $1 \leq j \leq n$ such that $x : A \in \Gamma_j$ and $\Gamma'_j, x : A, \Gamma''_j \vdash \sum_{i=1}^{i_j}N_{ji}$ and by IH we have a derivation
    \[
    (\sum_{i=1}^{i_j}N_{ji})[M_1 + \ldots + M_m / x] = \sum_{i=1}^{i_j}\sum_{k=1}^{m}N_{ji}[M_k / x]
    \]
    \begin{gather*}
    \Gamma'_j,\Gamma,\Gamma''_j \vdash (N_{j 1} + \ldots + N_{j i_j})[M_1 + \ldots + M_m / x] =\\
      = N_{j 1}[M_1 / x] + \ldots + N_{j 1}[M_m /x] + \ldots + N_{j i_j}[M_1 / x] + \ldots + N_{j i_j}[M_m /x]
    \end{gather*}
    and by applying the rule above we have
    \begin{gather*}
     \Gamma_1, \ldots, \Gamma'_j,\Gamma,\Gamma''_j, \ldots, \Gamma_n \vdash \sum_{k_1=1}^{i_1} \ldots \sum_{k_j=1}^{i_j}\sum_{l=1}^{m} \ldots \sum_{k_n=1}^{i_n} f(N_{1k_1}, \ldots, N_{jk_j}[M_{l} / x], \ldots, N_{nk_n}) =\\
     = \sum_{k_1=1}^{i_1}\ldots\sum_{k_n=1}^{i_n}f(N_{1k_1}, \ldots, N_{j,k_j}[M_1 + \ldots + M_m], \ldots, N_{nk_n})\\
     = \sum_{k_1=1}^{i_1}\ldots\sum_{k_n=1}^{i_n}f(N_{1k_1}, \ldots, N_{nk_n})[M_1 + \ldots + M_m]
    \end{gather*}
  \item Suppose $N$ is the conclusion of the rule
  \[
    \inferrule*[right=$\otimes I$]{\Delta_1 \vdash \sum_{i=1}^{p}P_i : A \quad \Delta_2 \vdash \sum_{i=1}^{n} N_i : B}{\Delta_1, \Delta_2 \vdash \sum_{i=1}^{p}\sum_{j=1}^{n}\{P_i, M_j\} : A \otimes B}
  \]
  and w.l.o.g suppose $x : A \in \Delta_1$, then by IH we have a derivation 
  \[
    \Delta'_1, \Gamma, \Delta''_1 \vdash (\sum_{i=1}^{p}P_i)[M_1 + \ldots + M_m] = (\sum_{i=1}^{p} \sum_{j=1}^{m}P_i[M_j / x]) : A
  \]
  and by applying the same rule we get
  \begin{gather*}
    \Delta'_1,\Gamma,\Delta''_1,\Delta_2 \vdash \sum_{i=1}^{p}\sum_{k=1}^{m}\sum_{j=1}^{n}\{P_i[M_j / x],N_j\} = \sum_{i=1}^{p}\sum_{j=1}^{n}\{P_i[M_1 + \ldots + M_m / x],N_j\} = \\
    = \sum_{i=1}^{p}\sum_{j=1}^{n}\{P_i,N_j\}[M_1 + \ldots + M_m / x]
  \end{gather*}
  \item Suppose $N$ is the conclusion of the rule
  \[
    \inferrule*[right=$+I$]{\Delta', x : A, \Delta'' \vdash \sum_{i=1}^{i_1} N_{1i} \quad \ldots \quad \Delta', x : A, \Delta'' \vdash \sum_{i=1}^{i_n} N_{ni}}{\Delta', x : A, \Delta'' \vdash \sum_{i=1}^{n}\sum_{j=1}^{i_j} N_{ij}}
  \]
  then by IH we have 
  \begin{gather*}
  \Delta',\Gamma,\Delta'' \vdash (\sum_{i=1}^{i_1}N_{1i})[M_1 + \ldots + M_m] = (\sum_{i=1}^{i_1}N_{1i}[M_1 + \ldots + M_m]) = (\sum_{i=1}^{i_1}\sum_{j=1}^{m}N_{1i}[M_j / x])\\
  \vdots\\
  \Delta',\Gamma,\Delta'' \vdash (\sum_{i=1}^{i_n}N_{ni})[M_1 + \ldots + M_m] = (\sum_{i=1}^{i_1}N_{ni}[M_1 + \ldots + M_m]) = (\sum_{i=1}^{i_1}\sum_{j=1}^{m}N_{ni}[M_j / x])\\
  \end{gather*}
  and again by applying the same rule we get
  \begin{gather*}
    \sum_{k=1}^{n}\sum_{i=1}^{i_k}\sum_{j=1}^{m}(N_{ki}[M_j / x]) = \sum_{k=1}^{n}\sum_{i=1}^{i_k}(N_{ki}[M_1 + \ldots + M_m / x]) = \sum_{k=1}^{n}\sum_{i=1}^{i_k}(N_{ki})[M_1 + \ldots + M_m / x]
  \end{gather*}
  \end{enumerate}
  The rest of the cases are analogous.
  \textcolor{red}{Some metastatement that substitution is robust with respect to the ambiguity in derivations?}
\end{proof}

\begin{lemma}[Substitution is associative and interchanging]
  TODO
\end{lemma}

\begin{example}
  Recall the raw morphism representation for our e-graph example
  \[(a \otimes ((2;\comonoid) \otimes (1;id));\sym);((* \otimes \counit + (id \otimes \counit \otimes id);<\!\!<) \otimes id);/~,\]
  We can render it using the  \enquote{enhanced} term syntax.
  \[
  \inferrule*{
  \inferrule*{
  \inferrule*{\vdash a : Int \quad \vdash 2 : Int}{\vdash a * 2}
  \quad
  \inferrule*{\vdash a : Int \quad \vdash 1 : Int}{\vdash a <\!\!< 1}
  }{\vdash (a * 2) + (a <\!\!< 1)}
  \quad
  \inferrule*{}{\vdash 2 : Int}
  }
  {\vdash ((a * 2) + (a <\!\!< 1)) / 2}
  \]
  Recall the string diagram
  \[
  \tikzfig{./figures/e-graph-example-string}
  \]
  We can linerise it as
  \begin{minted}{haskell}
    let (x_1, x_2, x_3) = (a, 2, 1) in
      let y_1 = {(x_1 * x_2), (x_1 << x_3)} in
        y1 / x_2
  \end{minted}
  \textcolor{red}{Why would one be preferable over the other? One syntax provides a reasoning framework about monoidal e-graphs while the other is just a linearisation of hypergraphs? Equality procedure for hypergraphs?}
\end{example}

>>>>

\usepackage{stmaryrd}
\newcommand{\parr}{\mathbin{\bindnasrepma}}

@@

\input{../preamble}
\input{../macros}
\input{../sample.tikzstyles}
\input{../hypergraph.tikzstyles}
\input{../hypergraph.tikzdefs}
\renewcommand*{\itemautorefname}{}

@@

In more detail, below is two different flavours of conjunction rules in a logic where all structural rules are allowed.

\[
\inferrule*[right=$\land L$]{\Gamma, A \vdash \Delta}{\Gamma, A \land B \vdash \Delta} \qquad \inferrule*[right=$\land R$]{\Gamma, B \vdash \Delta}{\Gamma, A \land B \vdash \Delta} \qquad \inferrule*[]{\Gamma \vdash A, \Delta \quad \Gamma \vdash B, \Delta}{\Gamma \vdash A \land B, \Delta}
\]
which is an \textit{additive} version, and
\[
\inferrule*[right=$\land$]{\Gamma, A, B \vdash \Delta}{\Gamma, A \land B \vdash \Delta} \qquad \inferrule*[]{\Gamma_1 \vdash A, \Delta_1 \quad \Gamma_2 \vdash B, \Delta_2}{\Gamma_1, \Gamma_2 \vdash A \land B, \Delta_1, \Delta_2}
\]
which is a \textit{multiplicative} version.

These presentations are equivalent in the presence of structural rules. For example, we can derive multiplicative $\land$ using the additive $\land L$ and $\land R$:

\[
  \inferrule{\Gamma, A, B \vdash \Delta}
  {\inferrule{\Gamma, B, A \land B \vdash \Delta}
    {\inferrule{
      \Gamma, A \land B, A \land B \vdash \Delta
      }{
        \Gamma, A \land B \vdash \Delta
        }}}
\]
and the other direction
\[
\inferrule[]{\Gamma, A \vdash \Delta}{
  \inferrule{\Gamma, A, B \vdash \Delta}{\Gamma, A \land B \vdash \Delta}}
\]

If we remove contraction and weakening, then the presentations are no longer equivalent and hence define different connectives, $\otimes$ and $\&$, respectively called \enquote{tensor} and \enquote{with} and also two connectives for multiplicative and additive disjunction $\parr$ and $\oplus$.

\[
\begin{array}{cc}
\textbf{Identity rules} & \inferrule*[right=$id$]{\;}{A, A^{\bot}} \qquad \inferrule*[right=cut]{\Gamma,A \quad A^{\bot},\Delta}{\Gamma, \Delta}\\
\textbf{Multiplicative rules} &  \inferrule*[right=$\parr$]{\Gamma, A, B}{\Gamma, A \parr B} \qquad \inferrule*[right=$\otimes$]{\Gamma,A \quad \Delta,B}{\Gamma,\Delta,A \otimes B}\\
\textbf{Additive rules} & \inferrule*[right=$\oplus_0$]{\Gamma,A}{\Gamma, A \oplus B} \qquad \inferrule*[right=$\oplus_1$]{\Gamma,B}{\Gamma, A \oplus B} \qquad \inferrule*[right=$\&$]{\Gamma,A \quad \Gamma, B}{\Gamma, A \& B}
\end{array}
\]  
Here's a useful intuition behind $\otimes$ and $\parr$. Suppose you are given two resources, a \textit{saw} and a \textit{hammer}, then you can construct $\text{saw} \otimes \text{hammer}$ which is a set of a saw and a hammer and you can give a saw to one person and the hammer to another.
In contrast, $\text{saw} \parr \text{hammer}$ is like a multitool, you can give it to a single person only.
The difference between $\&$ and $\oplus$ is a bit more subtle. 
Consider the menu below

\begin{minipage}{0.45\linewidth}
  \begin{tabular}{c}
    \textbf{Menu (price 17 Euros)}\\
    \textit{Quiche or Salad}\\
    \textit{Chicken or Fish}\\
    \textit{Banana or \enquote{Surprise du Chef}}\textsubscript{*}\\
    {}\\
    \textit{(*) either \enquote{Profiteroles} or \enquote{Tarte Tatin}}

  \end{tabular}
\end{minipage}
\hfill
\begin{minipage}{0.45\linewidth}
  \[
  \begin{aligned}
  17E \vdash 
  \left\{ \begin{array}{c}
    (Q \& S)\\
    \otimes\\
    (C \& F)\\
    \otimes\\
    (B \& (P \oplus T))
  \end{array}
  \right.
\end{aligned}
  \]
\end{minipage}

So let us start again from the beginning, considering a play between the restaurant manager (the Player) and the customer (the Opponent). 
It is the Player's responsibility to split the $17E$ into three parts, corresponding to the cost of the three parts of the meal. 
May be, this is done as follows:
\[
\inferrule*{5E \vdash Q \& S \quad 8E \vdash C \& F \quad 4E \vdash B \& (P \oplus T)}{17E \vdash (Q \& S) \otimes (C \& F) \otimes (B \& (P \oplus T))}
\]
Now let the Opponent challenge $5E \vdash Q \& S$:
\[
\inferrule*[]{5E \vdash Q \qquad 5E \vdash S}{5E \vdash Q \& S}
\]

which reads as: both Quiche and Salad are available to the customer, but he can get only one, and it is his choice of picking one of the antecedents and to order, say, a Quiche. 
Thus the additive conjunction can be understood as a $\ldots$ disjunction embodying a notion of external choice (remember that in our example the customer is the Opponent, or the context, or the environment). 
Let us now analyse a proof of $4E \vdash B \& (P \otimes T)$:

\[
\inferrule*{4E \vdash B  \qquad \inferrule*[]{4E \vdash T}{4E \vdash P \oplus T}}{4E \vdash B \& (P \oplus T)}
\]

Suppose that the Opponent chooses the Surprise. 
Then it is the Player's turn, who justifies $4E \vdash O \oplus T$ using the right $\oplus$ rule. 
So, the Opponent will get a Tarte Tatin, but the choice was in the Player's hands. 
Thus $\oplus$ has an associated meaning of internal choice. 
In summary, two forms of choice, external and internal, are modelled by N and $\oplus$, respectively. 
In the case of $\oplus$, whether $A$ or $B$ is chosen is controlled by the rule, that is, by the Player. 
In the case of $\&$, the choice between $A$ or $B$ is a choice of one of the antecedents of the rule, and is in the hands of the Opponent.

All these oppositions confirm a fundamental polarity: by convention, we shall term $\&$ and $\parr$ as negative, and $\otimes$ and $\oplus$ as positive.

\begin{remark}

Linear logic does not have the property that terms uniquely determine derivations.
Consider the following two derivations for example
\[
\inferrule*[right=$id$]{\;}{A \oplus B, A^{\bot} \;\&\; B^{\bot}} \qquad \inferrule*[]{\inferrule*[]{\inferrule*[right=$id$]{\;}{A,A^{\bot}}}{A \oplus B, A^{\bot}} \quad \inferrule*[]{\inferrule*[right=$id$]{\;}{B,B^{\bot}}}{A \oplus B, B^{\bot}}}{A \oplus B, A^{\bot} \;\&\; B^{\bot}}
\]

as negation is defined as
\begin{gather*}
(A \;\&\; B)^{\bot} = A^{\bot} \oplus B^{\bot}\\
(A \otimes B)^{\bot} = A^{\bot} \parr B^{\bot}
\end{gather*}

Therefore, such derivations need to be identified.
Quotienting by the identity above and a bunch of others defines $\equiv_{\eta}$.
Quotienting by \texttt{cut}  defines $\equiv_\beta$.
\end{remark}

\begin{remark}
We also have that some rules $r_1,r_2$ commute in the sense that they can be applied in any order without loss of provability.
For example,
\[
\inferrule*[right=$\&$]{\inferrule*[right=$\parr$]{\Gamma, A, B, C \quad \Gamma, A, B, D}{\Gamma, A, B, C \;\&\; D}}{\Gamma, A \parr B, C \;\&\; D} 
\qquad
\inferrule*[right=$\&$]{\inferrule*[right=$\parr$]{\Gamma, A, B, C}{\Gamma, A \parr B, C} \quad \inferrule*[right=$\parr$]{\Gamma, A, B, D}{\Gamma, A \parr B, D}}{\Gamma, A \parr B, C \;\&\;D}
\]
\end{remark}

@@

\begin{lemma}[Substitution is associative and interchanging]

  \begin{enumerate}
  \item If $\Gamma \vdash M : A$ and $\Delta, x : A, \Delta' \vdash N : B$ and $\Psi, y : B, \Psi' \vdash P : C$, then
  \[
  (P[M/x])[N/y] = P[M[N/y]/x]
  \]
  \item If $\Gamma \vdash M : A$ and $\Delta \vdash N : B$ and $\Psi, x : A, \Psi', y : B, \Psi'' \vdash P : C$, then
  \[
  (P[M/x][N/y]) = P[N/y][M/x]
  \]
  \end{enumerate}
\end{lemma}

@@

\begin{lemma}[Substitution]
  If $\Gamma \vdash M_1 + \ldots + M_m : A$ and $\Delta, x : A, \Phi \vdash N$ are derivable, then $\Delta, \Gamma, \Phi \vdash N[M/x]$ is derivable.
\end{lemma}
\begin{proof}
  Since the substitution is admissible in the original type theory, we only need to consider the extra rule that we introduced.
  Suppose $N \Coloneqq N_1 + \ldots + N_n$ and therefore comes from the rule
  \[
  \inferrule[]{\inferrule[]{\vdots}{\Gamma \vdash M} \quad \inferrule[]{\Delta, x : A, \Phi \vdash N_1, \ldots, \Delta, x : A, \Phi \vdash N_n}{\Delta, x : A, \Phi \vdash N}}{\;}
  \]
  By IH we have derivations
  \[
    \Delta, \Gamma, \Phi \vdash N_1[M_1 + \ldots + M_m /x], \ldots, \Delta, \Gamma, \Phi \vdash N_n[M_1 + \ldots + M_m /x]
  \]
  or,
  \[
    \Delta, \Gamma, \Phi \vdash N_1[M_1/ x] + \ldots + N_1[M_m /x], \ldots, \Delta, \Gamma, \Phi \vdash N_n[M_1 / x] + \ldots + N_n[M_m /x]
  \]
  and therefore we can get a derivation 
  \[
  \Delta, \Gamma, \Phi \vdash (N_1[M_1/ x] + \ldots + N_1[M_m /x]) + \ldots + (N_n[M_1/ x] + \ldots + N_n[M_m /x])
  \]

@@

Consider the revised rules (forget about the symmetry for a moment) in \autoref{fig:slat-typetheory}.
\begin{figure}
\begin{gather*}
  % \inferrule*[right=$id$]{\vdash A\;\text{type}}{x : A \vdash x + \ldots + x : A}\\
  \inferrule*[right=$id$]{\vdash A\;\text{type}}{x : A}\\
  \inferrule*[right=$f$]{f \in \mathcal{G}(A_1,\ldots,A_n;B) \quad \Gamma_1 \vdash \sum_{j=1}^{i_1} M_{1j} : A_1 \ldots \Gamma_n \vdash \sum_{j=1}^{i_n} M_{nj} : A_n}{\Gamma_1, \ldots, \Gamma_n \vdash \sum_{k_1 = 1}^{j_1} \ldots \sum_{k_n=1}^{i_n} f(M_{1 k_1}, \ldots, M_{n k_n}) : B}\\
  \inferrule*[right=$\otimes I$]{\Gamma \vdash \sum_{i=1}^{m}M_i : A \quad \Delta \vdash \sum_{i=1}^{n} N_i : B}{\Gamma, \Delta \vdash \sum_{i=1}^{m}\sum_{j=1}^{n}\{M_i, M_j\} : A \otimes B}\\
  \inferrule*[right=$\otimes E$]{\Gamma \vdash \sum_{i=1}^{m}M_i : A \otimes B \qquad \Delta_1, x : A, y : B, \Delta_2 \vdash \sum_{i=1}^{n}N_i : C}{\Delta_1, \Gamma, \Delta_2 \vdash \sum_{1}^{i=m}\sum_{j=1}^{n}\text{match}_{A \otimes B}(M_i, xy. N_j) : C}\\
  \inferrule*[right=$\mathbf{1}I$]{\;}{() \vdash \star : \mathbf{1}}\\
  \inferrule*[right=$\mathbf{1}E$]{\Gamma \vdash \sum_{i=1}^{m} M_i : \mathbf{1} \qquad \Delta_1,\Delta_2 \vdash \sum_{i=1}^{n} N_i : A}{\Delta_1,\Gamma,\Delta_2 \vdash \sum_{i=1}^{m}\sum_{j=1}^{n} \text{match}_{\mathbf{1}}(M_i,N_j) : A}\\
  \inferrule*[right=$+I$]{\Gamma \vdash \sum_{i=1}^{i_1} M_{1i} : A \quad \ldots \quad \Gamma \vdash \sum_{i=1}^{i_m} M_{mi} : A}{\Gamma \vdash \sum_{i=1}^{m}\sum_{j=1}^{i_j} M_{ij} : A}
\end{gather*}
\caption{Type theory for semilattice-enriched monoidal categories}
\label{fig:slat-typetheory}
\end{figure}

@@

\begin{definition}[Substitution]
  Given derivations $\Gamma \vdash M_1 + \ldots + M_m : A$ and $\Delta_1, x : A, \Delta_2 \vdash N$,
  we define $\Delta_1, \Gamma, \Delta_2 \vdash N[M_1 + \ldots + M_m / x]$ by induction on derivation of $N$.
  Moreover, we have the property that $(N_1 + \ldots + N_n)[M_1 + \ldots + M_m / x] = \sum_{i=1}^{n}\sum_{j=1}^{m}N_i[M_j / x]$.
  \begin{enumerate}
    \item Suppose $N \Coloneqq x : A \vdash x : A$, then $A = B$ and we can just use $M_1 + \ldots + M_m$ and define $\Gamma \vdash x[M_1 + \ldots + M_m / x] = x[M_1 / x] + \ldots + x[M_m / x] = M_1 + \ldots + M_m$.
    \item Suppose $N$ is the conclusion of the rule
    \[
      \inferrule*[right=$fI$]{f \in \mathcal{G}(A_1,\ldots,A_n;B) \quad \Gamma_1 \vdash \sum_{j=1}^{i_1} M_{1j} : A_1 \ldots \Gamma_n \vdash \sum_{j=1}^{i_n} M_{nj} : A_n}{\Gamma_1, \ldots, \Gamma_n \vdash \sum_{k_1 = 1}^{j_1} \ldots \sum_{k_n=1}^{i_n} f(M_{1 k_1}, \ldots, M_{n k_n}) : B}
    \] 
    then there exists $1 \leq j \leq n$ such that $x : A \in \Gamma_j$ and $\Gamma'_j, x : A, \Gamma''_j \vdash \sum_{i=1}^{i_j}N_{ji}$ and by IH we have a derivation
    \[
    (\sum_{i=1}^{i_j}N_{ji})[M_1 + \ldots + M_m / x] = \sum_{i=1}^{i_j}\sum_{k=1}^{m}N_{ji}[M_k / x]
    \]
    \begin{gather*}
    \Gamma'_j,\Gamma,\Gamma''_j \vdash (N_{j 1} + \ldots + N_{j i_j})[M_1 + \ldots + M_m / x] =\\
      = N_{j 1}[M_1 / x] + \ldots + N_{j 1}[M_m /x] + \ldots + N_{j i_j}[M_1 / x] + \ldots + N_{j i_j}[M_m /x]
    \end{gather*}
    and by applying the rule above we have
    \begin{gather*}
     \Gamma_1, \ldots, \Gamma'_j,\Gamma,\Gamma''_j, \ldots, \Gamma_n \vdash \sum_{k_1=1}^{i_1} \ldots \sum_{k_j=1}^{i_j}\sum_{l=1}^{m} \ldots \sum_{k_n=1}^{i_n} f(N_{1k_1}, \ldots, N_{jk_j}[M_{l} / x], \ldots, N_{nk_n}) =\\
     = \sum_{k_1=1}^{i_1}\ldots\sum_{k_n=1}^{i_n}f(N_{1k_1}, \ldots, N_{j,k_j}[M_1 + \ldots + M_m], \ldots, N_{nk_n})\\
     = \sum_{k_1=1}^{i_1}\ldots\sum_{k_n=1}^{i_n}f(N_{1k_1}, \ldots, N_{nk_n})[M_1 + \ldots + M_m]
    \end{gather*}
  \item Suppose $N$ is the conclusion of the rule
  \[
    \inferrule*[right=$\otimes I$]{\Delta_1 \vdash \sum_{i=1}^{p}P_i : A \quad \Delta_2 \vdash \sum_{i=1}^{n} N_i : B}{\Delta_1, \Delta_2 \vdash \sum_{i=1}^{p}\sum_{j=1}^{n}\{P_i, M_j\} : A \otimes B}
  \]
  and w.l.o.g suppose $x : A \in \Delta_1$, then by IH we have a derivation 
  \[
    \Delta'_1, \Gamma, \Delta''_1 \vdash (\sum_{i=1}^{p}P_i)[M_1 + \ldots + M_m] = (\sum_{i=1}^{p} \sum_{j=1}^{m}P_i[M_j / x]) : A
  \]
  and by applying the same rule we get
  \begin{gather*}
    \Delta'_1,\Gamma,\Delta''_1,\Delta_2 \vdash \sum_{i=1}^{p}\sum_{k=1}^{m}\sum_{j=1}^{n}\{P_i[M_j / x],N_j\} = \sum_{i=1}^{p}\sum_{j=1}^{n}\{P_i[M_1 + \ldots + M_m / x],N_j\} = \\
    = \sum_{i=1}^{p}\sum_{j=1}^{n}\{P_i,N_j\}[M_1 + \ldots + M_m / x]
  \end{gather*}
  \item Suppose $N$ is the conclusion of the rule
  \[
    \inferrule*[right=$+I$]{\Delta', x : A, \Delta'' \vdash \sum_{i=1}^{i_1} N_{1i} \quad \ldots \quad \Delta', x : A, \Delta'' \vdash \sum_{i=1}^{i_n} N_{ni}}{\Delta', x : A, \Delta'' \vdash \sum_{i=1}^{n}\sum_{j=1}^{i_j} N_{ij}}
  \]
  then by IH we have 
  \begin{gather*}
  \Delta',\Gamma,\Delta'' \vdash (\sum_{i=1}^{i_1}N_{1i})[M_1 + \ldots + M_m] = (\sum_{i=1}^{i_1}N_{1i}[M_1 + \ldots + M_m]) = (\sum_{i=1}^{i_1}\sum_{j=1}^{m}N_{1i}[M_j / x])\\
  \vdots\\
  \Delta',\Gamma,\Delta'' \vdash (\sum_{i=1}^{i_n}N_{ni})[M_1 + \ldots + M_m] = (\sum_{i=1}^{i_1}N_{ni}[M_1 + \ldots + M_m]) = (\sum_{i=1}^{i_1}\sum_{j=1}^{m}N_{ni}[M_j / x])\\
  \end{gather*}
  and again by applying the same rule we get
  \begin{gather*}
    \sum_{k=1}^{n}\sum_{i=1}^{i_k}\sum_{j=1}^{m}(N_{ki}[M_j / x]) = \sum_{k=1}^{n}\sum_{i=1}^{i_k}(N_{ki}[M_1 + \ldots + M_m / x]) = \sum_{k=1}^{n}\sum_{i=1}^{i_k}(N_{ki})[M_1 + \ldots + M_m / x]
  \end{gather*}
  \end{enumerate}
  The rest of the cases are analogous.
\end{definition}

Then the admissibility follows from the above construction.
\begin{lemma}
  If we have derivations $\Gamma \vdash M_1 + \ldots + M_n : A$ and $\Delta_1, x : A, \Delta_2 \vdash N : B$, then we can derive $\Delta_1, \Gamma, \Delta_2 \vdash N[M_1 + \ldots + M_m / x]$.
\end{lemma}

The ambiguity technically means that we can not recursively define the action of a substitution on terms, however, we can see that applying the substitution to different derivations produces the same result
\[
  \inferrule*[right=$fI$]{f \in \mathcal{G}(A;B) \qquad \inferrule*[right=$+I$]{x : A \vdash x[M_1 + M_2 / x] = M_1 + M_2 : A \quad x : A \vdash x[M_1 + M_2 / x] = M_1 + M_2 : A}{x : A \vdash (x + x)[M_1 + M_2 / x] = (M_1 + M_2) + (M_1 + M_2) \equiv M_1 + M_2 + M_1 + M_2 : A}}{x : A \vdash (f(x) + f(x))[M_1 + M_2 / x] = f(M_1 + M_2) + f(M_1 + M_2) \equiv f(M_1) + f(M_2) + f(M_1) + f(M_2) : B}
\]
\[
  \inferrule*[right=$+I$]{\inferrule*[right=$fI$]{f \in \mathcal{G}(A;B) \quad x : A \vdash x[M_1 + M_2 / x] = M_1 + M_2 : A}{x : A \vdash f(x)[M_1 + M_2 / x] = f(M_1) + f(M_2)} \quad \inferrule*[right=$fI$]{f \in \mathcal{G}(A;B) \quad x : A \vdash x[M_1 + M_2 / x] = M_1 + M_2 : A}{x : A \vdash f(x)[M_1 + M_2 / x] = f(M_1) + f(M_2)}}{x : A \vdash (f(x) + f(x))[M_1 + M_2 / x] = f(M_1) + f(M_2) + f(M_1) + f(M_2) : B}
\]
\begin{lemma}[Substitution is associative and interchanging]
  We have the following properties for the substitution.
  \begin{enumerate}[label=(\alph*)]
    \item\label{subst_prop:a} $(N_1 + \ldots + N_n)[M_1 + \ldots + M_m / x] = \sum_{i=1}^{n}\sum_{j=1}^{m}N_i[M_j / x]$ (shown above)
    \item\label{subst_prop:b} If $\Gamma \vdash N : B$ and $\Delta, y : B, \Delta' \vdash M : A$ and $\Psi, x : A, \Psi' \vdash P : C$, then
      \[
      (P[M/x])[N/y] = P[M[N/y]/x]
      \]
    \item\label{subst_prop:c} If $\Gamma \vdash M : A$ and $\Delta \vdash N : B$ and $\Psi, x : A, \Psi', y : B, \Psi'' \vdash P : C$, then
      \[
      (P[M/x][N/y]) = P[N/y][M/x]
      \]
  \end{enumerate}
\begin{proof}
  \leavevmode\\
  We will first prove~\autoref{subst_prop:b}.
  \begin{itemize}
    \item Suppose $P \Coloneqq x : A \vdash x : A$, then the result of $P[M/x] = M$ and therefore $(P[M/x])[N / y] = P[M[N/y]/x] = M[N/y]$
    \item Suppose 
    \begin{gather*}
      P \Coloneqq \inferrule[]{
        f \in \mathcal{G}(A_1, \ldots, A_n;C) \quad \Gamma_1 \vdash \sum_{j=1}^{i_1} M_{1j} : A_1 \ldots \Gamma_n \vdash \sum_{j=1}^{i_n}M_{nj} : A_n
        }{
          \Gamma_1, \ldots, \Gamma_n \vdash \sum_{k_1=1}^{j_1}\ldots\sum_{k_n=1}^{i_n}f(M_{1k_1},\ldots,M_{nk_{n}}): C
        }
    \end{gather*}
    then 
    \[
